Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-06-01 article/main.tex
section 6

@@

We ran it on a server equipped with two 16-core \textsf{Intel Xeon Gold 6130 CPU
  @ 2.10GHz} ``\emph{Skylake}'' CPUs. The inner loop does $2^{37}$ iterations
and terminates in 42.3s, which makes 23 core minutes.

@@

When the increment $c$ is known, the internal state of \textsf{PCG64} can be
practically reconstructed from $X_0, \dots, X_{42}$ using the algorithms shows
in section~\ref{sec:Cunknown}. Only algorithm~\ref{algo:unknown_1} is
computationally expensive; we implemented it in \textsf{C}, while we implemented
algorithms~\ref{algo:unknown_2} and~\ref{algo:last} in \textsf{Python}.

We have shown that algorithm~\ref{algo:unknown_1} is correct when $\ell=13$. In
fact, we empirically observed that it returns the right result with probability
$\approx 66\%$ when $\ell=11$ and with probability $99\%$ when
$\ell=12$. Decrementing $\ell$ reduces the workload by a factor of 4. Therefore,
if $T$ denotes the running time when $\ell=13$, we actually used the following
strategy:
\begin{algorithmic}[1]
\State Run with $\ell=11$ (if it succeeds, stop).
\State Run with $\ell=11$ with different inputs.
\State Run with $\ell=11$ with different inputs.
\State Run with $\ell=12$.
\State Run with $\ell=13$.
\end{algorithmic}
It is not difficult to check that the expected running time of this procedure is
$\frac{1079}{10800} T \approx 0.1 T$. In our implementation, $T = 200,000$ CPU
hours, so the expected running time of the reconstruction procedure is about
$20,000$ CPU hours. In fact we were lucky: on the challenge input, the first
attempt with $\ell=11$ succeeded and took only $12,500$ CPU hours.

The whole process actually ran in 35 wall-clock minutes using 512 cluster nodes,
each equipped with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz}
(``\emph{Cascade Lake}''). The actual machine is the
\verb|[----REDACTED FOR PEER REVIEW----]| computer located at the
\verb|[----ALSO REDACTED----]| computation center. Note that on this particular
parallel computer, running the algorithm with $\ell=13$ would take 10 hours
using the same amount of resources, so the whole procedure \emph{is} practical,
even in the absolute worst case.

The outer loop of algorithm~\ref{algo:unknown_1} makes $2^{2\ell - 1}$
iterations while the inner loop makes $2^{30}$ iterations. Using a single
hardware execution context, we measured that one of the outer loop takes between
41.5s and 44s (apparently not all nodes of the cluster are running at exactly
the same speed, potentially because of ``turbo boost'' adjustments and thermal
constraints; most of the time, they take exactly... 42 seconds). Because of this
variability, we implemented a master-slave work distribution pattern, in which a
master process dispatches iterations of the outer loop to slave processes. This
also made checkpointing very easy. We used \textsf{MPI} for inter-process
communication.

>>>>

The analysis of section~\ref{sec:geometric} yields parameters that guarantee
that the reconstruction procedure \emph{always} succeeds. In most cases, these
parameters are pessimistic. We ran a series of experiments to determine more
practical choices: using smaller-than-guaranteed values of $\ell$ (the number of
guessed least-significant bits), we measured the success probability of the
state reconstruction procedure. The results are shown in table~\ref{tab:empirical}.

\begin{table}
  \centering
\begin{tabular}{|c|c|c|c|c|}
  \cline{1-2}\cline{4-5}
  \multicolumn{2}{|c|}{$n=3$ (section~\ref{sec:Cknown})} & &  \multicolumn{2}{c|}{$n=4$ (section~\ref{sec:cvp_small_2})} \\
  \cline{1-2}\cline{4-5}
  $\ell$ & Success proba. && $\ell$ & Success proba. \\
  \cline{1-2}\cline{4-5}
  \noalign{\vskip\doublerulesep
  \vskip-\arrayrulewidth}
  \cline{1-2}\cline{4-5}
  16 & $\approx 0.125$     && 10 &  $\approx 0.12$ \\
  \cline{1-2}\cline{4-5}
  17 & $\approx 0.25$     && 11 &  $\approx 0.64$ \\
  \cline{1-2}\cline{4-5}
  18 & $\approx 0.5$      && 12 &  $\approx 0.995$ \\
  \cline{1-2}\cline{4-5}
  19 & 1                  && 13 &  1 \\
  \cline{1-2}\cline{4-5}
  20 & 1 (proved)         && 14 &  1 (proved) \\
  \cline{1-2}\cline{4-5}
\end{tabular}
\caption{Empirical success probabilities with smaller parameters.}
\label{tab:empirical}
\end{table}

@@

From section~\ref{sec:cvp_small}, we know that guessing $\ell=20$
least-significant bits ensures deterministic success. However, we observed
empirically that $\ell=19$ works with probability $\approx 1$, and runs twice as
fast. $\ell=18$ and $\ell=17$ run with probability $\approx 1/2$ and
$\approx 1/4$ respectively, therefore are much less useful. In practice, we used
$\ell=19$.

We ran it on a server equipped with two 16-core \textsf{Intel Xeon Gold 6130 CPU
  @ 2.10GHz} (``\emph{Skylake}'') CPUs. The inner loop does $2^{37}$ iterations
and terminates in 42.3s, which makes 23 core minutes.

@@

When the increment $c$ is known, the internal state of \textsf{PCG64} can be
practically reconstructed from $X_0, \dots, X_{63}$ using the algorithms shown
in section~\ref{sec:Cunknown}. Only algorithm~\ref{algo:unknown_1} is
computationally expensive; we implemented it in \textsf{C}, while we implemented
algorithms~\ref{algo:unknown_2} and~\ref{algo:last} in \textsf{Python}.

We have shown that algorithm~\ref{algo:unknown_1} is correct when $\ell=14$. The
procedure does $2^{29 + 2\ell}$ iterations of the inner loop, so decreasing
$\ell$ would really be interesting. Looking at table~\ref{tab:empirical}, we
settle for $\ell=13$ in the worst case; let $T$ denotes the running time when
$\ell=13$.

It seems that the most promising strategy consists in choosing $\ell=11$; if the
reconstruction procedure fails, then we try again with different inputs. The
expected running time of this approach number of trials is
$T/(16 \times 0.64) \approx T / 10.25$. In our implementation, $T = 200,000$ CPU
hours, so the expected running time of the reconstruction procedure is about
$20,000$ CPU hours. In fact we were lucky: on the challenge input, the first
attempt with $\ell=11$ succeeded, so the whole process took only $12,500$ CPU
hours.

It actually ran in 35 wall-clock minutes using 512 cluster nodes, each equipped
with two 20-cores \textsf{Intel Xeon Gold 6248 @ 2.5Ghz} (``\emph{Cascade
  Lake}''). The actual machine is the \verb|[----REDACTED FOR PEER REVIEW----]|
computer located at the \verb|[----ALSO REDACTED----]| computation center. Note
that on this particular parallel computer, running the algorithm with $\ell=13$
would take 10 hours using the same amount of resources, so the whole procedure
\emph{is} practical, even in the absolute worst case.

The outer loop of algorithm~\ref{algo:unknown_1} makes $2^{2\ell - 1}$
iterations while the inner loop makes $2^{30}$ iterations. Using a single
hardware execution context, we measured that one of the outer loop takes between
41.5s and 44s (apparently not all nodes of the cluster are running at exactly
the same speed, potentially because of ``turbo boost'' adjustments and thermal
constraints). Because of this variability, we implemented a master-slave work
distribution pattern, in which a master process dispatches iterations of the
outer loop to slave processes. This also made checkpointing very easy. We used
\textsf{MPI} for inter-process communication.
